Bayesian spectral density estimation using P-spline priors

### 2020-03-27 pSplines.tex
updated results
Updated results for the simulation study

\textcolor{red}{For non-equidistant knots, it does not make sense any longer to base the definition of the covariance matrix of the Gaussian prior  on the  difference matrices as defined in (\ref{eq:D1}) and (\ref{eq:D2}) because this is an adequate approximation of the derivative-based roughness measures -- defined as the integral of the first and second order derivatives of the P-splines -- 
for equally spaced knots only. The definition of the difference matrices could be adjusted using divided differences, but here we use the derivative-based penalties as described by \cite{WoodSimon2017Pwdb}. We make use of the implementation of the derivative-based penalty matrix  by the function {\tt bsplinepen} in the R-package {\tt fda}. This penalty matrix contains the inner products of the respective first and second order derivatives of the basis functions.}

@@

The P-spline analysis was performed using the \texttt{gibbs\_pspline} function implemented in the \textsf{R} package \texttt{psplinePsd}.  A total of 100,000 MCMC samples were generated
using  a pilot run of 25,000 iterations with a burn-in period of 5,000 to calibrate the proposals.  The final sample consisted of 75,000 samples with a burn-in period of 5,000 and thinning factor of 10, resulting in 7,000 samples used for posterior inferences.  
%Therefore, the number of iterations equate the B-spline analysis.  
This procedure was replicated for both equidistant and \textcolor{red}{adaptive} knots.

@@

%REMARK: PLEASE INSERT GRAPH WITH AR(1) AND AR(4) PSD ESTIMATES
\def\arraystretch{1.1}
% IAE
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &  0.825 & 0.769 & 0.609  \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ &  0.729 & 0.632 & 0.517  \\
			\hspace{0.5em}P-spline $d=2$&  0.844 & 0.612 & 0.638  \\
			\hspace{0.5em}P-spline $d=3$&  0.727 & 0.728 & 0.769  \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 0.910 & 0.769 & 0.595 \\
			\hspace{0.5em}P-spline $d=2$ & 0.994 & 0.867 & 0.731 \\
			\hspace{0.5em}P-spline $d=3$ & 1.059 & 0.829 & 0.610 \\ \hline
			AR(4)     &  &  &  \\ 
			B-spline & 3.019 & 2.250 & 1.832   \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 3.014 & 2.462 & 2.335  \\
			\hspace{0.5em}P-spline $d=2$& 3.247 & 2.619 & 2.399 \\
			\hspace{0.5em}P-spline $d=3$& 3.438 & 2.786 & 2.510  \\	
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 2.659 & 1.884 & 1.465 \\
			\hspace{0.5em}P-spline $d=2$ & 2.627 & 1.849 & 1.411 \\
			\hspace{0.5em}P-spline $d=3$ & 2.625 & 1.835 & 1.410 \\
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Median IAE for the psd estimates using the BspDP and the P-spline priors for simulated AR(1) and AR(4) time series with  different penalty orders ($d$) and knot location schemes for the P-splines.}
	\label{table:sim_IAE}
\end{table}

@@

For the AR(1) model, the BspDP and first order penalty P-spline 90\% credible bands cover the whole true spectral density function, i.e.\ for all frequencies $\lambda$, in all the 300 independent analyses.  In general, the latter method shows diminishing  performance as the penalty degree increases.  For the AR(4) model, the P-spline method with equidistant knots performs poorly.  On the other hand, the \textcolor{red}{adaptive} knot scheme yields evidently much better results, which are comparable (in the case of the first order penalty) to those obtained via the BspDP algorithm.  For $n=\{128,256\}$, the latter has higher coverage probabilities than our proposal, except for $n=512$.  For the P-spline priors, the coverage probabilities decrease as the penalty degree order increases, which can be explained by the forced smoothness imposed by the higher penalty order.  As shown by \cite{Edwards2019}, the BspDP algorithm produces better results than the BDP prior in the AR(4) case, which coverage probability is zero across $n=\{128,256,512\}$. Consequently, our P-spline algorithm with \textcolor{red}{adaptive} knots outperforms the BDP method in this case.

@@

% Coverage probability Table
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &  1.000 & 1.000 & 1.000 \\
			\textit{Equidistant}& & &   \\
			\hspace{0.5em}P-spline $d=1$& 1.000 & 1.000 & 1.000  \\
			\hspace{0.5em}P-spline $d=2$& 0.987 & 0.993 & 1.000  \\
			\hspace{0.5em}P-spline $d=3$& 0.943 & 0.877 & 0.870  \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 1.000 & 1.000 & 1.000 \\
			\hspace{0.5em}P-spline $d=2$ & 0.993 & 0.980 & 0.990 \\
			\hspace{0.5em}P-spline $d=3$ & 0.947 & 0.840 & 0.957 \\ \hline		       
			AR(4)     &  &  &  \\ 
			B-spline &  0.933 & 0.967 & 0.897  \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 0.400 & 0.353 & 0.010 \\
			\hspace{0.5em}P-spline $d=2$ & 0.200 & 0.167 & 0.003 \\
			\hspace{0.5em}P-spline $d=3$ & 0.053 & 0.023 & 0.000 \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\	
			\hspace{0.5em}P-spline $d=1$ & 0.820 & 0.887 & 0.920 \\
			\hspace{0.5em}P-spline $d=2$ & 0.637 & 0.700 & 0.773 \\
			\hspace{0.5em}P-spline $d=3$ & 0.470 & 0.600 & 0.583 \\	
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Proportion of times that the true psd is entirely encapsulated within the 90\% uniform credible band of the 300 estimates.}
	\label{table:sim_covPbb}
\end{table}

\begin{table}
	\centering
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &  1.000 & 1.000 & 1.000 \\
			\textit{Equidistant \&}& & &   \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1,2,3$ & 1.000 & 1.000 & 1.000  \\ \hline	 		      
			AR(4)     &  &  &  \\ 
			B-spline &  1.000 & 1.000 & 1.000 \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 0.984 & 0.992 & 0.984 \\
			\hspace{0.5em}P-spline $d=2$& 0.984 & 0.984 & 0.980  \\
			\hspace{0.5em}P-spline $d=3$& 0.968 & 0.976 & 0.969  \\	
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\	
			\hspace{0.5em}P-spline $d=1$ & 1.000 &  1.000 &  1.000 \\
			\hspace{0.5em}P-spline $d=2$ & 1.000 &  1.000 &  1.000 \\
			\hspace{0.5em}P-spline $d=3$ & 0.984 &  1.000 &  1.000 \\		
			\bottomrule
		\end{tabular}
	\caption{Median of the pointwise  proportions that the true psd at each Fourier frequency is contained within the $90\%$ uniform credible bands.}
	\label{table:sim_covPbb_median}
\end{table}

@@

As \cite{Edwards2019} noted, one of the drawbacks of the BspDP prior  is its computational complexity relative to the BDP.  This is directly reflected in the run-time, which is approximately 2-3 times higher in this example (see Table~3 in \cite{Edwards2019}).  The median run-time in our analyses for the BspDP prior and the P-spline with \textcolor{red}{adaptive} knots are displayed in Table~\ref{table:sim_time}.  Note that the P-spline method based on equidistant knots has similar run-times to the \textcolor{red}{adapative} knots.  The P-spline method is approximately 3-6 times faster than BspDP approach in these examples.  This is due to a reduced number of calculations  per iteration because the B-spline densities are calculated only  once at the start of the algorithm and there are no Dirichlet process approximations via the stick-breaking method.

% TIME Table
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &   29.497 & 32.423 & 37.896 \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$& 5.629 & 8.291 & 10.507 \\
			\hspace{0.5em}P-spline $d=2$& 5.650 & 8.374 & 10.512 \\
			\hspace{0.5em}P-spline $d=3$& 5.587 & 8.301 & 10.427 \\ 
			B-spline/P-spline &  5.247 & 3.896 & 3.615 \\ \hline
			AR(4)     &  &  & \\ 
			B-spline & 30.111 & 34.126 & 40.875  \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 5.621 & 8.240 & 10.462 \\
			\hspace{0.5em}P-spline $d=2$& 5.641 & 8.372 & 10.549 \\
			\hspace{0.5em}P-spline $d=3$& 5.611 & 8.327 & 10.686 \\
			B-spline/P-spline & 5.354 & 4.105 & 3.869 \\	
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Median run-times in minutes and relative run-times with respect to an average of the P-spline run-times.}
	\label{table:sim_time}
\end{table}

@@

%Overall, the P-spline prior outperforms the B-spline prior in this example in terms of IAE, coverage probability and run-time.  Particularly, our novel knot allocation scheme, the periodogram-spaced knots, yield the best results in the case of peaked spectral densities.  The  rule-of-thumb criterion,  $K=\min\left\{n/4, 40\right\}$, for the number of 
\textcolor{red}{P-splines} %B-spline densities in the P-spline prior algorithm
 worked perfectly well in this example.

>>>>

\textcolor{red}{For non-equidistant knots, it does not make sense any longer to base the definition of the covariance matrix of the Gaussian prior  on the  difference matrices as defined in (\ref{eq:D1}) and (\ref{eq:D2}) because this is an adequate approximation of the derivative-based roughness measures -- defined as the integral of the first and second order derivatives of the P-splines -- 
for equally spaced knots only. The definition of the difference matrices could be adjusted using divided differences, but here we use the derivative-based penalties as described by \cite{WoodSimon2017Pwdb}. We make use of the implementation of the derivative-based penalty matrix  by the function {\tt bsplinepen} in the R-package {\tt fda}. This penalty matrix contains the inner products of the respective first and second order derivatives of the basis functions.} {\pmr In addition, we normalise this matrix by dividing it by the maximum absolute column sum.}

@@

The P-spline analysis was performed using the \texttt{gibbs\_pspline} function implemented in the \textsf{R} package \texttt{psplinePsd}.  A total of 100,000 MCMC samples were generated
using  a pilot run of 20,000 iterations with a burn-in period of 5,000 and thinning factor of 10 to calibrate the proposals.  The final sample consisted of 80,000 samples with a burn-in period of 5,000 and thinning factor of 10, resulting in 7,500 samples used for posterior inferences.  
%Therefore, the number of iterations equate the B-spline analysis.  
This procedure was replicated for both equidistant and \textcolor{red}{adaptive} knots{\pmr considering first and second order penalties.}

@@

%REMARK: PLEASE INSERT GRAPH WITH AR(1) AND AR(4) PSD ESTIMATES
\def\arraystretch{1.1}
% IAE
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &  0.825 & 0.769 & 0.609  \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ &  0.772 & 0.644 & 0.527 \\
			\hspace{0.5em}P-spline $d=2$&  0.708 & 0.604 & 0.629 \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 0.874 & 0.783 & 0.610 \\
			\hspace{0.5em}P-spline $d=2$ & 0.874 & 0.815 & 0.640 \\ \hline
			AR(4)     &  &  &  \\ 
			B-spline & 3.019 & 2.250 & 1.832   \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ &  2.914 & 2.389 & 2.307\\
			\hspace{0.5em}P-spline $d=2$&  3.150 & 2.569 & 2.388\\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 2.495 & 1.830 & 1.489\\
			\hspace{0.5em}P-spline $d=2$ &  2.696 & 2.003 & 1.677\\
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Median IAE for the psd estimates using the BspDP and the P-spline priors for simulated AR(1) and AR(4) time series with  different penalty orders ($d$) and knot location schemes for the P-splines.}
	\label{table:sim_IAE}
\end{table}

@@

For the AR(1) model, the BspDP and first order penalty P-spline 90\% credible bands cover the whole true spectral density function, i.e.\ for all frequencies $\lambda$, in {\pmr almost} all the 300 independent analyses.  In general, the latter method shows diminishing  performance as the penalty degree increases.  For the AR(4) model, the P-spline method with equidistant knots performs poorly.  On the other hand, the \textcolor{red}{adaptive} knot scheme yields evidently much better results, which are comparable (in the case of the first order penalty) to those obtained via the BspDP algorithm.  %For $n=\{128,256\}$, the latter has higher coverage probabilities than our proposal, except for $n=512$.  
For the P-spline priors, the coverage probabilities decrease as the penalty degree order increases, which can be explained by the forced smoothness imposed by the higher penalty order.  As shown by \cite{Edwards2019}, the BspDP algorithm produces better results than the BDP prior in the AR(4) case, which coverage probability is zero across $n=\{128,256,512\}$. Consequently, our P-spline algorithm with \textcolor{red}{adaptive} knots outperforms the BDP method in this case.

@@

% Coverage probability Table
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &  1.000 & 1.000 & 1.000 \\
			\textit{Equidistant}& & &   \\
			\hspace{0.5em}P-spline $d=1$&  1.000 & 1.000 & 0.997\\
			\hspace{0.5em}P-spline $d=2$& 0.993 & 1.000 & 0.993  \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 0.997 & 1.000 & 0.997 \\
			\hspace{0.5em}P-spline $d=2$ & 0.973 & 0.740 & 0.770 \\ \hline		       
			AR(4)     &  &  &  \\ 
			B-spline &  0.933 & 0.967 & 0.897  \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ &  0.527 & 0.353 & 0.007 \\
			\hspace{0.5em}P-spline $d=2$ & 0.323 & 0.167 & 0.007\\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\	
			\hspace{0.5em}P-spline $d=1$ & 0.790 & 0.847 & 0.813 \\
			\hspace{0.5em}P-spline $d=2$ & 0.337 & 0.387 & 0.270 \\
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Proportion of times that the true psd is entirely encapsulated within the 90\% uniform credible band of the 300 estimates.}
	\label{table:sim_covPbb}
\end{table}

\begin{table}
	\centering
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &  1.000 & 1.000 & 1.000 \\
			\textit{Equidistant \&}& & &   \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1,2$ & 1.000 & 1.000 & 1.000  \\ \hline	 		      
			AR(4)     &  &  &  \\ 
			B-spline &  1.000 & 1.000 & 1.000 \\
			\textit{Equidistant knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ &  1.000 & 0.992 & 0.984 \\
			\hspace{0.5em}P-spline $d=2$&  0.984 & 0.984 & 0.980 \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\	
			\hspace{0.5em}P-spline $d=1$ & 1.000 & 1.000 & 1.000 \\
			\hspace{0.5em}P-spline $d=2$ & 0.984 & 0.992 & 0.992\\	
			\bottomrule
		\end{tabular}
	\caption{Median of the pointwise  proportions that the true psd at each Fourier frequency is contained within the $90\%$ uniform credible bands.}
	\label{table:sim_covPbb_median}
\end{table}

@@

As \cite{Edwards2019} noted, one of the drawbacks of the BspDP prior  is its computational complexity relative to the BDP.  This is directly reflected in the run-time, which is approximately 2-3 times higher in this example (see Table~3 in \cite{Edwards2019}).  The median run-time in our analyses for the BspDP prior and the P-spline with \textcolor{red}{adaptive} knots are displayed in Table~\ref{table:sim_time}.  Note that the P-spline method based on equidistant knots has similar run-times to the \textcolor{red}{adapative} knots.  The P-spline method is approximately {\pmr 4-7} times faster than BspDP approach in these examples.  This is due to a reduced number of calculations  per iteration because the B-spline densities are calculated only  once at the start of the algorithm and there are no Dirichlet process approximations via the stick-breaking method.

% TIME Table
\begin{table}
	\centering
%	\begin{ruledtabular}
		\begin{tabular}{lccc}
			\toprule
			& $n=128$ & $n=256$ & $n=512$ \\ \hline
			AR(1)     &  &  &  \\ 
			B-spline &   29.497 & 32.423 & 37.896 \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$&  4.859 & 7.450 & 9.606\\
			\hspace{0.5em}P-spline $d=2$&  4.538 & 6.863 & 8.783\\
			B-spline/P-spline &  6.278 & 4.531 & 4.122 \\ \hline
			AR(4)     &  &  & \\ 
			B-spline & 30.111 & 34.126 & 40.875  \\
			\textit{\textcolor{red}{Adaptive} knots}& & &   \\
			\hspace{0.5em}P-spline $d=1$ & 4.883 & 7.453 & 9.665 \\
			\hspace{0.5em}P-spline $d=2$& 4.543 & 6.887 & 8.827 \\
			B-spline/P-spline & 6.389 & 4.760 & 4.421 \\	
			\bottomrule
		\end{tabular}
%	\end{ruledtabular}
	\caption{Median run-times in minutes and relative run-times with respect to an average of the P-spline run-times.}
	\label{table:sim_time}
\end{table}

@@

%Overall, the P-spline prior outperforms the B-spline prior in this example in terms of IAE, coverage probability and run-time.  Particularly, our novel knot allocation scheme, the periodogram-spaced knots, yield the best results in the case of peaked spectral densities.  The  rule-of-thumb criterion,  $K=\min\left\{n/4, 40\right\}$, for the number of 
%\textcolor{red}{P-splines} B-spline densities in the P-spline prior algorithm
% worked perfectly well in this example.
